### 2024-01-02 manuscript.Rmd
add zenodo release ref

authornote: |
  This work was conducted within the project “Lifespan Rhythms of Memory and Cognition (RHYME)” at the Max Planck Institute for Human Development. ESB was a fellow of the International Max Planck Research School on the Life Course (LIFE; http://www.imprs-life.mpg.de/en). AP was a fellow of the International Max Planck Research School on Computational Methods in Psychiatry and Ageing Research (COMP2PSYCH, https://www.mps-ucl-centre.mpg.de/comp2psych). MW-B received support from the Jacobs Foundation (Early Career Research Fellowship 2017–2019, 2016-1217-16). CTN’s work is supported by a grant from the German Research Foundation (DFG; NG 191/2-1) and an Early Career Research Fellowship by the Jacobs Foundation (2021-1417-99). We are grateful to all members of the RHYME and LIME projects for valuable feedback on the project. Finally, we acknowledge support by the Max Planck Dahlem Campus of Cognition (MPDCC). The authors declare no competing interests.
  The latest supplementary materials for this article are available at <https://github.com/ebuchberger/Estimating-Power-for-SEM>. The version of code and data used as the basis for this PDF is "`r paste0(repro::current_hash(backend = 'gert'))`".

@@

The model is defined by a model string, which specifies three latent factors of memory (Pattern Separation [ps]; Pattern Completion, [pc]; Generalization, [gen]) which are measured through the manifest variables (ps: tasks 1 to 6, pc: tasks 5 to 9, gen: tasks 10 to 13). Note that lavaan entails several user-friendly functions for fitting models that automatically include default settings for model components that are not explicitly specified (e.g. adding residual variances, covariances of exogenous latent variables). Thus, the model syntax can be kept very concise (check documentation of the respective function for full information on default settings).

Estimating statistical power for model comparisons requires that researchers specify each of the competing models and all aspects in which they differ from one another (e.g., in number of latent factors, allocations of manifest variables; see supplementary R code on GitHub for specifications of Models 1 and 2).

@@

Here, we define the generate function for the tripartite model. The factor loadings specified in step 2 (Design) are called from the condition argument such that they are varied across different iterations of the simulation, and are defined as "a" for syntax readability. For standardization purposes (that is, the total variance of each manifest variable summing up to 1), the loadings of indicators Task5 and Task6 are set to $\frac {a}{\sqrt{(2+2 \cdot cov1)}}$ to account for cross-loadings on multiple factors and are defined as "b" to ease syntax readability (cov1 = covariance between the latent factors pattern separation and pattern completion). Both loading parameters a and b are stored in the list "parameters" such that they can be called via the glue_data function from the glue package [@hester_2022] to concatenate the model string. The residual variances of the manifest variables are set to $1-a^2$ for standardization purposes. The population model is defined by a model string, which specifies three latent factors (ps, pc, gen) measured by the 13 manifest indicators (task1 – task13), the residual variances as defined above and the covariance between the latent factors as defined in the condition argument. The data are simulated from the model syntax via the simulateData() function from the lavaan package [@rosseel_2012] and stored as "dat".

@@

### 2.1.4 Step 4: Analyze
In this step, we define the analyze function, which specifies how the data generated in step 3 should be analyzed. For the comparison of non-nested models, we rely on a combination of relative and absolute model fit parameters (for discussion of an alternative model selection approach, see Section 4). We fit each of the competing models to each of the simulated data sets using the sem() function from the lavaan package [@rosseel_2012] with a maximum likelihood estimator and extract different measures of model fit using the helper function get_fitmeasures() (see funs.R in the supplementary material on GitHub), namely the Comparative Fit Index [CFI, @bentler_1990], root mean square error of approximation [RMSEA, @steiger_2016] and the Bayesian Information Criterion [BIC, @raftery_1986]. The function returns the fit measures, the simulation model that was used in the respective iteration and information on model convergence. The analyze function for the example on memory composition could thus look like this: 
```{r analyze, echo = TRUE}
```

@@

SEM offers a powerful tool for modeling multivariate relationships between psychological constructs. Maximizing its utility in psychological sciences hinges on the integration of its complex characteristics with power estimation – another key tenet of scientific rigor. Our work exemplifies a roadmap for navigating some of the challenges in harmonizing these two efforts by implementing a-priori power estimation within an SEM framework. In the case of information-theoretic model comparisons for non-nested models, statistical power cannot be determined analytically, i.e., model comparisons cannot be performed by using $\chi^2$-based statistics. Monte Carlo simulations provide a useful tool to circumvent this methodological challenge [@muthen_2002], allowing researchers to derive informed decisions on sample size planning to ensure sufficient statistical power and prevent uninformative or misleading research [@button_2013]. In this study, we illustrate a step-by-step approach to implementing a Monte Carlo simulation in the statistical programming language R to estimate the required sample size for model comparisons of non-nested models. The complete code is available online to yield a starting point for researchers who want to implement their own Monte Carlo power simulations (https://github.com/ebuchberger/Estimating-Power-for-SEM).

@@

While we demonstrate in this tutorial that simulation-based techniques help overcome methodological shortcomings from traditional analytical approaches, the randomness that is inherent to such approaches may seem at odds with efforts in results reproducibility and replicability. Here, a clear-cut solution is through transparent and accessible documentation of the simulation code and adequate software management. That is, researchers should make use of repositories on platforms such as GitHub or the Open Science Framework to make code scripts available to other researchers, and clearly specify the versions of software programs and packages that were used (e.g. via containerization or at least in written form). We thus urge researchers to seriously consider the issue of reproducibility for their simulation analyses, to make outcomes of such a-priori power estimates easily accessible to other researchers and reviewers [for a more detailed guideline on a reproducible workflow in R see @peikert_2021 and the implementation of this manuscript on GitHub as an example]. Note, however, that—specifically in cases when simulations are parallelized on computing clusters—the exact reproducibility of simulations as demonstrated in this tutorial might be compromised, even when a seed is specified in the analysis script. Nevertheless, a sufficiently high number of iterations ensures replicability, i.e., comparable estimates of statistical power. Derived conclusions about required sample size will therefore remain unaffected.

@@

  The supplementary materials for this article (including the data and R scripts) are publicly available at <https://github.com/ebuchberger/Estimating-Power-for-SEM>. The analyses reported in this manuscript were not preregistered.

>>>>

authornote: |
  This work was conducted within the project “Lifespan Rhythms of Memory and Cognition (RHYME)” at the Max Planck Institute for Human Development. ESB was a fellow of the International Max Planck Research School on the Life Course (LIFE; http://www.imprs-life.mpg.de/en). AP was a fellow of the International Max Planck Research School on Computational Methods in Psychiatry and Ageing Research (COMP2PSYCH, https://www.mps-ucl-centre.mpg.de/comp2psych). MW-B received support from the Jacobs Foundation (Early Career Research Fellowship 2017–2019, 2016-1217-16). CTN’s work is supported by a grant from the German Research Foundation (DFG; NG 191/2-1) and an Early Career Research Fellowship by the Jacobs Foundation (2021-1417-99). We are grateful to all members of the RHYME and LIME projects for valuable feedback on the project. Finally, we acknowledge support by the Max Planck Dahlem Campus of Cognition (MPDCC). The authors declare no competing interests.
  The latest supplementary materials for this article are available on [GitHub](https://github.com/ebuchberger/Estimating-Power-for-SEM) [@zenodo-release]. The version of code and data used as the basis for this PDF is "`r paste0(repro::current_hash(backend = 'gert'))`".

@@

The model is defined by a model string, which specifies three latent factors of memory (Pattern Separation [ps]; Pattern Completion, [pc]; Generalization, [gen]) which are measured through the manifest variables (tasks 1 to 13). Note that lavaan entails several user-friendly functions for fitting models that automatically include default settings for model components that are not explicitly specified (e.g. adding residual variances, covariances of exogenous latent variables). Thus, the model syntax can be kept very concise (check documentation of the respective function for full information on default settings).

Estimating statistical power for model comparisons requires that researchers specify each of the competing models and all aspects in which they differ from one another [e.g., in number of latent factors, allocations of manifest variables\; For specifications of Models 1 and 2 see supplementary R code on [GitHub](https://github.com/ebuchberger/Estimating-Power-for-SEM), @zenodo-release].

@@

Here, we define the generate function for the tripartite model. The factor loadings specified in step 2 (Design) are called from the condition argument such that they are varied across different iterations of the simulation, and are defined as "a" for syntax readability. For standardization purposes (that is, the total variance of each manifest variable summing up to 1), the loadings of indicators Task5 and Task6 are set to $\frac {a}{\sqrt{(2+2 \cdot cov1)}}$ to account for cross-loadings on multiple factors and are defined as "b" to ease syntax readability (cov1 = covariance between the latent factors pattern separation and pattern completion). Both loading parameters a and b are stored in the list "parameters" such that they can be called via the glue_data function from the glue package [@hester_2022] to concatenate the model string. The residual variances of the manifest variables are set to $1-a^2$ for standardization purposes. The population model is defined by a model string, which specifies three latent factors (ps, pc, gen) measured by the 13 manifest indicators (ps: tasks 1 to 6, pc: tasks 5 to 9, gen: tasks 10 to 13), the residual variances as defined above and the covariance between the latent factors as defined in the condition argument. The data are simulated from the model syntax via the simulateData() function from the lavaan package [@rosseel_2012] and stored as "dat".

@@

### 2.1.4 Step 4: Analyze
In this step, we define the analyze function, which specifies how the data generated in step 3 should be analyzed. For the comparison of non-nested models, we rely on a combination of relative and absolute model fit parameters (for discussion of an alternative model selection approach, see Section 4). We fit each of the competing models to each of the simulated data sets using the sem() function from the lavaan package [@rosseel_2012] with a maximum likelihood estimator and extract different measures of model fit using the helper function get_fitmeasures() [see funs.R in the supplementary material on [GitHub](https://github.com/ebuchberger/Estimating-Power-for-SEM), @zenodo-release], namely the Comparative Fit Index [CFI, @bentler_1990], root mean square error of approximation [RMSEA, @steiger_2016] and the Bayesian Information Criterion [BIC, @raftery_1986]. The function returns the fit measures, the simulation model that was used in the respective iteration and information on model convergence. The analyze function for the example on memory composition could thus look like this: 
```{r analyze, echo = TRUE}
```

@@

SEM offers a powerful tool for modeling multivariate relationships between psychological constructs. Maximizing its utility in psychological sciences hinges on the integration of its complex characteristics with power estimation – another key tenet of scientific rigor. Our work exemplifies a roadmap for navigating some of the challenges in harmonizing these two efforts by implementing a-priori power estimation within an SEM framework. In the case of information-theoretic model comparisons for non-nested models, statistical power cannot be determined analytically, i.e., model comparisons cannot be performed by using $\chi^2$-based statistics. Monte Carlo simulations provide a useful tool to circumvent this methodological challenge [@muthen_2002], allowing researchers to derive informed decisions on sample size planning to ensure sufficient statistical power and prevent uninformative or misleading research [@button_2013]. In this study, we illustrate a step-by-step approach to implementing a Monte Carlo simulation in the statistical programming language R to estimate the required sample size for model comparisons of non-nested models. The complete code is available on [GitHub](https://github.com/ebuchberger/Estimating-Power-for-SEM) [@zenodo-release] to yield a starting point for researchers who want to implement their own Monte Carlo power simulations.

@@

While we demonstrate in this tutorial that simulation-based techniques help overcome methodological shortcomings from traditional analytical approaches, the randomness that is inherent to such approaches may seem at odds with efforts in results reproducibility and replicability. Here, a clear-cut solution is through transparent and accessible documentation of the simulation code and adequate software management. That is, researchers should make use of repositories on platforms such as GitHub or the Open Science Framework to make code scripts available to other researchers, and clearly specify the versions of software programs and packages that were used (e.g. via containerization or at least in written form). We thus urge researchers to seriously consider the issue of reproducibility for their simulation analyses, to make outcomes of such a-priori power estimates easily accessible to other researchers and reviewers [for a more detailed guideline on a reproducible workflow in R see @peikert_2021 and the implementation of this manuscript on [GitHub](https://github.com/ebuchberger/Estimating-Power-for-SEM) as an example]. Note, however, that—specifically in cases when simulations are parallelized on computing clusters—the exact reproducibility of simulations as demonstrated in this tutorial might be compromised, even when a seed is specified in the analysis script. Nevertheless, a sufficiently high number of iterations ensures replicability, i.e., comparable estimates of statistical power. Derived conclusions about required sample size will therefore remain unaffected.

@@

  The supplementary materials for this article (including the data and R scripts) are publicly available on [GitHub](https://github.com/ebuchberger/Estimating-Power-for-SEM) [@zenodo-release]. The analyses reported in this manuscript were not preregistered.
